Quantifying the ease of viral escape from broad and narrow antibodies to influenza hemagglutinin

### 2017-10-22 paper/manuscript.tex
work on neut assay section of results

\begin{figure}[h!]
\centerline{\includegraphics[width=0.91\textwidth]{figs/logoplots_pymol/logoplots_pymol.pdf}}
\caption{
\label{fig:structures}
{\bf Mutations selected by broad and narrow antibodies.}
(A) The logo plots show the sites where mutations have the largest effect.
Letter heights are proportional to the excess fraction of virions with that mutation that survive antibody, as indicated by the scale bars.
The structures are colored white to red by the excess fraction surviving for the largest-effect mutation at each site, with each antibody scaled separately.  
(B) Sites of selection from anti-stalk antibodies, with the same coloring scale for both antibodies.
(C) Cladogram of HA subtypes, with group 1 in black and group 2 in gray. 
The amino acid at site 38 is indicated. 
Colors indicate whether a subtype is bound or neutralized by C179 if this has been tested.  
}
\end{figure}

For the broad anti-receptor binding pocket antibody S139/1, we observe strong escape mutations at sites 156, 158, and 193 (Figure~\ref{fig:structures}A; sites are in H3 numbering). 
These three sites fall directly in the physical binding footprint of the antibody~\citep{lee2012heterosubtypic}.
These are also the same three sites where \citet{yoshida2009cross} selected escape mutants in H1, H2, and H3 HAs. 
Our data show that numerous different amino-acid mutations at each site confer neutralization resistance.
The mutation with the largest effect, D158N, introduces an N-linked glycosylation motif.

Although anti-stalk antibodies C179 and FI6v3 only select mutations with small effects, these mutations almost all fall in the physical binding footprints of the antibodies (Figure~\ref{fig:structures}A).
The two antibodies have similar epitopes and angles of approach~\citep{dreyfus2013structure}, and they select identical mutations at several sites (Figure~\ref{fig:structures}B). 
The three largest-effect mutations for FI6v3 (K280S, K280T, and N291S) all introduce N-linked glycosylation motifs near the epitope, and all three mutations also have a similar effect size for C179.

However, C179 selects several mutations that do not have any apparent effect on FI6v3 (Figure~\ref{fig:structures}A, Figure~\ref{suppfig:FI6v3logo}).
The most notable of these C179-specific mutations occurs at site 38.
The additional breadth of FI6v3 over antibodies such as C179 that neutralize only group 1 HAs has been attributed to FI6v3's ability to accommodate a glycan on the asparagine at site 38 that is present in group 2 but not group 1 HAs~\citep{corti2011neutralizing,sui2009structural,ekiert2009antibody}. 
However, the H38S mutation that has the largest effect on C179 resistance in our experiments does not introduce a glycosylation motif, showing that there are also other ways to escape anti-stalk antibodies at this site.
Interestingly, group 1 HA subtypes that are susceptible by C179 tend to possess a histidine at site 38, but subtypes that are not bound or neutralized by C179 often possess a serine (Figure~\ref{fig:structures}C). 
%The amino-acid identity at site 111 of HA2 can change the orientation of a conserved Trp21 in HA2, also resulting in group 1 and group 2 differences in the binding ability of stalk-targeting antibodies. An H111T mutation in the A/South Carolina/1/1918 (H1N1) has been reported to abrogate C179 binding~\citep{dreyfus2013structure}.
%However, we did not observe escape mutations at site 111, but this may be due to strain-specific differences.

Interestingly, FI6v3 also weakly selects several mutations at residue -8, which is part of HA's signal peptide (Figure~\ref{fig:structures}A). 
This signal peptide is cleaved from the mature HA protein~\citep{daniels2003n,burke2014recommended}, so presumably these mutations are affecting the density of HA trimers on the virion~\citep{nordholm2017translational} rather than the physical binding of the antibody to HA.

\comment{still need to add something about signal peptide mutation, maybe refer to H17* antibodies too}

\subsection*{Validating escape mutation effect sizes with neutralization assays}

\comment{pdmH1N1 has an NTT glycosylation motif at sites 278-280. Its IC50 for FI6v3 is around 4-fold higher than the IC50 for WSN, according to Corti et al 2011}

@@

>>>>

\begin{figure}[h!]
\centerline{\includegraphics[width=0.89\textwidth]{figs/logoplots_pymol/logoplots_pymol.pdf}}
\caption{
\label{fig:structures}
{\bf Mutations selected by broad and narrow antibodies.}
(A) Logo plots show sites where mutations have the largest effect.
Letter heights are proportional to the excess fraction of virions with that mutation that survive antibody, as indicated by the scale bars.
Structures are colored white to red by the excess fraction surviving for the largest-effect mutation at each site, with each antibody scaled separately.  
(B) Sites of selection from anti-stalk antibodies, with the same coloring scale for both antibodies.
(C) Cladogram of HA subtypes, with group 1 in black and group 2 in gray. 
The amino acid at site 38 is indicated. 
Colors indicate whether a subtype is bound or neutralized by C179.  
}
\end{figure}

For the broad anti-receptor binding pocket antibody S139/1, we observe strong escape mutations at sites 156, 158, and 193 (Figure~\ref{fig:structures}A; sites are in H3 numbering). 
These three sites fall directly in the physical binding footprint of the antibody~\citep{lee2012heterosubtypic}, and are the same three sites where \citet{yoshida2009cross} selected escape mutants in H1, H2, and H3 HAs. 
Our data show that numerous different amino-acid mutations at each site confer neutralization resistance.
The mutation with the largest effect, D158N, introduces an N-linked glycosylation motif.

Although the anti-stalk antibodies C179 and FI6v3 only select mutations with small effects, these mutations almost all fall in the physical binding footprints of the antibodies (Figure~\ref{fig:structures}A).
The two antibodies have similar epitopes and angles of approach~\citep{dreyfus2013structure}, and they select identical mutations at several sites (Figure~\ref{fig:structures}B). 
The three largest-effect mutations for FI6v3 (K280S, K280T, and N291S) all introduce glycosylation motifs near the epitope, and all three mutations have similar magnitude antigenic effects in both FI6v3 and C179.

However, C179 selects several mutations that do not have any apparent effect on FI6v3 (Figure~\ref{fig:structures}A, Figure~\ref{suppfig:FI6v3logo}).
The most notable of these C179-specific mutations occur at site 38.
The additional breadth of FI6v3 over antibodies such as C179 that neutralize only group 1 HAs is because FI6v3 can accommodate a glycan on the asparagine at site 38 that is present in group 2 HAs~\citep{corti2011neutralizing,sui2009structural,ekiert2009antibody}. 
However, the H38S mutation that has the largest effect on C179 resistance in our experiments does not introduce a glycosylation motif, showing that there are also other ways to escape anti-stalk antibodies at this site.
Interestingly, group 1 HA subtypes that are susceptible by C179 tend to possess a histidine at site 38, but subtypes that are not bound or neutralized by C179 often possess a serine (Figure~\ref{fig:structures}C). 
%The amino-acid identity at site 111 of HA2 can change the orientation of a conserved Trp21 in HA2, also resulting in group 1 and group 2 differences in the binding ability of stalk-targeting antibodies. An H111T mutation in the A/South Carolina/1/1918 (H1N1) has been reported to abrogate C179 binding~\citep{dreyfus2013structure}.
%However, we did not observe escape mutations at site 111, but this may be due to strain-specific differences.

Interestingly, FI6v3 weakly selects several mutations at residue -8, which is part of HA's signal peptide (Figure~\ref{fig:structures}A). 
This signal peptide is cleaved from the mature HA protein~\citep{daniels2003n,burke2014recommended}, so mutations at this site may affect the density of HA trimers on the virion~\citep{nordholm2017translational} rather than the physical binding of the antibody to HA.

\subsection*{Neutralization assays validate the mutational antigenic profiling}
Do the mutations identified in our mutational antigenic profiling actually have the expected effect on antibody neutralization?
We have previously validated many of the large-effect antigenic mutations selected by the narrow antibodies H17-L19, H17-L10, and H17-L7~\citep{doud2017complete}.
However, the mutations selected by the broad anti-stalk antibodies have much smaller effects -- especially for the broadest antibody, FI6v3.
We therefore tested some of these FI6v3-selected mutations using neutralization assays on individual viral mutants.

@@

Figure~\ref{fig:FI6v3neutcurves} shows that the mutational antigenic profiling is highly predictive of the results of the neutralization assays, even for small-effect mutations.
As discussed in the previous section, the three mutations most strongly selected by FI6v3 introduce glycosylation motifs at sites 278-280 or 289-291 (Figure~\ref{fig:structures}A).
We created viruses carrying each of these mutations (K280S, K280T, and N291S) and validated that all three mutants were modestly more resistant to FI6v3 (Figure~\ref{fig:FI6v3neutcurves}A).
As a control, we also validated that a mutation at one of these sites (K280A) that does \emph{not} have an effect in our mutational antigenic profiling does not shift the neutralization curve (Figure~\ref{fig:FI6v3neutcurves}A).
Interestingly, the HA of pandemic H1N1 has a glycosylation motif at 278-280, and \citet{corti2011neutralizing} have reported that this HA is $\sim$4-fold more resistant to FI6v3 than the A/WSN/1933 HA used in our experiments.
This reported difference between strains is of similar magnitude to the effect that we measure for introducing the glycosylation motif via the K280S or K280T mutations.

Our mutational antigenic profiling also identified several non-glycosylation-motif mutations that were selected by FI6v3.
We validated that one of these mutations, G47R in the HA2 chain, indeed increased neutralization resistance (Figure~\ref{fig:FI6v3neutcurves}A) -- although as predicted by the mutational antigenic profiling, the magnitude of the effect was smaller than for the glycosylation-motif mutations.
The most unexpected mutations identified in the mutational antigenic profiling were at site -8 in the signal peptide.
We tested one of these mutations, K(-8)T, and found that it indeed caused a modest but detectable increase in neutralization resistance (Figure~\ref{fig:FI6v3neutcurves}A).
As controls, we also tested three more mutations (P80D, V135T, and M17L in HA2) that did \emph{not} have substantial effects in the mutational antigenic profiling, and confirmed that none of them affected neutralization resistance.

A remarkable aspect of these validation experiments is the very small effect sizes of the identified mutations.
Antigenic mutations selected by strain-specific antibodies to HA generally increase the concentration of antibody needed to neutralize the virus by orders of magnitude.
Neutralization curves for a few examples of such large-effect escape mutants are in Figure~\ref{fig:FI6v3neutcurves}B,C.
Our work demonstrates that there are no such large-effect single mutations to the A/WSN/1933 HA when it comes to escaping from FI6v3 and C179.
But the results in Figure~\ref{fig:FI6v3neutcurves}A show that we can still leverage the high sensitivity of mutational antigenic profiling to identify mutations that have much more modest but nonetheless measurable effects on antibody neutralization.
